Bootstrap learning via modular concept discovery

### 2013-01-29 drafts/ijcai2013/ijcai13.tex
Chnages.

Let a \emph{task} be a function $t : \mathcal{L} \rightarrow \{0,
1\}$, where $\mathcal{L}$ is the set of all expressions in our
representation language. We say that $e \in \mathcal{L}$ \emph{solves}
task $t$ if $t(e) = 1$. Our goal is to solve as many tasks as
possible, as efficiently as possible. That is, suppose that there is a
distribution over tasks $t \sim  p_t(\cdot)$, we would like an ordering $O
= e_1, \dots $ on expressions in $\mathcal{L}$ such that if $p_t(t_1)
> p_t(t_2)$ then the first expression in $O$ to solve $t_1$ comes
before the first expression in $O$ to solve $t_2$.

We motivate a solution to this problem by considering a heirarchical
Bayesian formulation of the task distribution. Suppose we believe that
there is some underlying structure to $p_t(\cdot)$ that is explained by
the fact that for each task $t$ there is a latent expressions $e \in
\mathcal{L}$ drawn from a stochastic grammar $\mathcal{G}$ and that

\begin{align}
e &\sim p_\mathcal{L}(\cdot) = \mathcal{G}\\
t | e &\sim \text{U}\{ \text{tasks solved by }e\}.
\end{align}

Then $p_t(t) = \sum_{e_j \in \mathcal{L} \text{ s.t. } t(e_j) = 1}
p_\mathcal{L}(e_j)$. If we make the assumption that for the set $E$ of
expressions in this sum, there is some expression $e^*$ such that
$p_\mathcal{L}(e^*) >> p_\mathcal{L}(e), e \neq e^*$, then we have, approximately, 
\[ p_t(t) \propto} p_\mathcal{L}(e^*). \]

This suggests a way to find the ordering $O$: infer a stochastic
grammar $G$, under the assumption that $G$ favors one solution to a
given task over any other, and order the elements of $\mathcal{L}$
according to $G$.

Thus, our basic procedure will be to search for a single ``good''
solution for as many tasks as we can, and to use those solutions to
estimate a stochastic grammar. Our search will be performed in the
order of the most recently inferred grammar. So as our ordering
becomes more efficient and we can solve more tasks, we learn better
stochastic grammars.

@@

Based on the motivation set out above, we present an iterative
strategy. Given a stochastic grammar $\mathcal{G}$ which defines a
distribution over the primitives from $\mathcal{L}$ and a integer
\emph{frontier size} $N$, we perform to steps in each iteration.  In
the first step, we explore the $N$ highest probability programs
according to $\mathcal{G}$. In the second step, we compress the
programs that solve any of our tasks, and we use that compression to
define a new grammar. Repeating these two steps, we effectively
reparameterize our search space to include programs that are more likely
to be solutions to our tasks.

@@

\begin{figure}[h]
    \begin{tikzpicture}[scale=.9,
        transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {}
      child {node [above=.3cm, text width=2cm] (a) {$(Int \rightarrow Int) \rightarrow Int 
          \rightarrow Int$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$Int \rightarrow Int$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
\end{figure}

@@

We constructed two task sets. The first of these was constructed
explicitly to contain familiar module structure. In this set of tasks,
we sampled $1000$ boolean circuits using AND, OR, and NOT gates. To
accomplish this, we first sampled either 1, 2, 3 inputs, then between
1 and 5 gates, randomly wiring the inputs of each new gate to the
output of one of the existing gates in the circuit. We continued this
sampling procedure until we had 1000 ``connected'' circuits, i.e.,
circuits all of whose output are wired to an input except for the last
one (the output gate). This resulted in $1000$ tasks consisting of
$82$ unique boolean functions, with a distribution of tasks as shown
in Figure~\ref{fig:boolhist}. In Figure~\ref{fig:boolLearningCurves1},
We present learning curves for frontier sizes 100, 500 and 10000,
using the same procedure as in Section~\ref{sec:symreg}.

@@

What kind of solutions does our algorithm find to common boolean
circuits? We might hope that the circuits it finds are ``simple.'' One
way to judge this is to compare the circuits it finds with the
circuits found in digital electronics textbooks. For example, the XOR
circuit can be drawn as a circuit with only NAND gates as in
Figure~\ref{fig:xor_circuit}. Our algorithm returns the solution:

\begin{alltt}
\small{
((S (B C (((B (C B nand)) S) 
(((B (C B nand)) S) (B C 
((B (B nand)) nand)))))) I)}. 
\end{alltt}
This looks more complicated. It should be understood, though, that the
routing combinators S, B, and C play the role of wiring. Note that
this solution has four NAND gates, the same number of as in the
standard circuit. Another interesting feature of this solution is that
it is not the smallest expression that is equivalent (modulo
reduction), since we can further reduce it. However, this reduction
would remove the two uses of \texttt{\small{(B (C B nand) S)}}. This is
a reused component that appears within the most highly weighted 20
combinators in the algorithm's induced grammar. We can view it
schematically with a circuit diagram, but it is not a circuit
fragment. Rather, it is a function that takes in a circuit of type
$bool \rightarrow bool \rightarrow bool$ and returns a circuit of type $bool
\rightarrow bool \rightarrow bool$. Figure~\ref{fig:component} shows
this component schematically.

@@

\subsection{Discussion}

>>>>

%% Let a \emph{task} be a function $t : \mathcal{L} \rightarrow \{0,
%% 1\}$, where $\mathcal{L}$ is the set of all expressions in our
%% representation language. We say that $e \in \mathcal{L}$ \emph{solves}
%% task $t$ if $t(e) = 1$. Our goal is to solve as many tasks as
%% possible, as efficiently as possible. That is, suppose that there is a
%% distribution over tasks $t \sim  p_t(\cdot)$, we would like an ordering $O
%% = e_1, \dots $ on expressions in $\mathcal{L}$ such that if $p_t(t_1)
%% > p_t(t_2)$ then the first expression in $O$ to solve $t_1$ comes
%% before the first expression in $O$ to solve $t_2$.

%% We motivate a solution to this problem by considering a heirarchical
%% Bayesian formulation of the task distribution. Suppose we believe that
%% there is some underlying structure to $p_t(\cdot)$ that is explained by
%% the fact that for each task $t$ there is a latent expressions $e \in
%% \mathcal{L}$ drawn from a stochastic grammar $\mathcal{G}$ and that

%% \begin{align}
%% e &\sim p_\mathcal{L}(\cdot) = \mathcal{G}\\
%% t | e &\sim \text{U}\{ \text{tasks solved by }e\}.
%% \end{align}

%% Then $p_t(t) = \sum_{e_j \in \mathcal{L} \text{ s.t. } t(e_j) = 1}
%% p_\mathcal{L}(e_j)$. If we make the assumption that for the set $E$ of
%% expressions in this sum, there is some expression $e^*$ such that
%% $p_\mathcal{L}(e^*) >> p_\mathcal{L}(e), e \neq e^*$, then we have, approximately, 
%% \[ p_t(t) \propto} p_\mathcal{L}(e^*). \]

%% This suggests a way to find the ordering $O$: infer a stochastic
%% grammar $G$, under the assumption that $G$ favors one solution to a
%% given task over any other, and order the elements of $\mathcal{L}$
%% according to $G$.

%% Thus, our basic procedure will be to search for a single ``good''
%% solution for as many tasks as we can, and to use those solutions to
%% estimate a stochastic grammar. Our search will be performed in the
%% order of the most recently inferred grammar. So as our ordering
%% becomes more efficient and we can solve more tasks, we learn better
%% stochastic grammars.

@@

Based on the motivation set out above, we present an iterative
strategy. Given a stochastic grammar $\mathcal{G}$ which defines a
distribution over the primitives from $\mathcal{L}$ and a integer
\emph{frontier size} $N$, we perform two steps in each iteration.  In
the first step, we explore the $N$ highest probability programs
according to $\mathcal{G}$. In the second step, we compress the
programs that solve any of our tasks, and we use that compression to
define a new grammar. Repeating these two steps, we effectively
reparameterize our search space to include programs that are more likely
to be solutions to our tasks.

@@

%% \begin{figure}[h]
%%     \begin{tikzpicture}[scale=.9,
%%         transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}
%%         ]
%%       \node [font=\scriptsize] (top) {}
%%       child {node [above=.3cm, text width=2cm] (a) {$(Int \rightarrow Int) \rightarrow Int 
%%           \rightarrow Int$}
%%         child {node [left=.5cm, below=1cm, text width=3cm] 
%%           {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
%%             \rightarrow \\ t_2 \rightarrow t_0$}}
%%         child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
%%       }
%%       child {node [below right=.3cm and 1cm] (b) {$Int \rightarrow Int$}
%%         child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
%%       };
%%     \end{tikzpicture}
%% \end{figure}

@@

We constructed two task sets. The first of these was constructed
explicitly to contain familiar module structure. In this set of tasks,
we sampled $1000$ boolean circuits using AND, OR, and NOT gates. To
accomplish this, we first sampled either 1, 2, 3 inputs, then between
1 and 5 gates, randomly wiring the inputs of each new gate to the
output of one of the existing gates in the circuit. We continued this
sampling procedure until we had 1000 ``connected'' circuits, i.e.,
circuits all of whose output are wired to an input except for the last
one (the output gate). This resulted in $1000$ tasks consisting of
$82$ unique boolean functions, with a distribution of tasks as shown
in Figure~\ref{fig:booldistr}. In Figure~\ref{fig:boolLearningCurves1},
We present learning curves for frontier sizes 100, 500 and 10000,
using the same procedure as in Section~\ref{sec:symreg}.

\begin{figure}
\includegraphics[width=\linewidth]{figures/circuitDistrComp.pdf}
\label{fig:booldistr}
\caption{Distribution of boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates; b) the
  first 1000 expressions from the grammar over programs \emph{before}
  learning; c) the first 1000 expressions from the grammar over
  programs \emph{after} learning.}
\end{figure}

@@

@@

\subsection{Discussion}
This work attempts to bring together ideas from various areas of
research -- hierarchical Bayes, learning to learn, minimum description
length learning -- to provide a proof of concept that learning
abstractions from programs via compression can rapidly improve search
performance. We emphasize this by focusing on domains in which the
solution space is not smooth and the error signal is all or
none. Thus, the main source of information for learning is the modular
and compositional structure of the solution space.

Of course, the representations that people use to learn in the world
often have a partially smooth or local structure. Techniques in AI and
machine learning commonly use smoothness or local structure to make
search tractable. But the difference between this kind of learning and
conceptual learning is undeniable. The former is what the statistician
or machine learning researcher implements. The latter is what she
learns as a student and as a practicioner. As researchers, we try
different models, watch them fail or succeed, and abstract out the
elements of these models that capture the relevant differences. We mix
and match parts of different ideas, and utilize succcessful patterns
of reasoning that can span many problems and domains.

We believe that the primary characteristic of conceptual learning is
that it occurs over a modular and compositional computational
representation that supports higher-order functions. That is why we
chose a functional programming language as the representational
language of this work.

The experiments in this work do not contain the richness of real world
learning, but we believe that the learning algorithm's behavior
exhibits some of the characteristics we would expect of a conceptual
learner.

One characteristic of this algorithm's behavior is a rapid initial
increase in performance followed by a hard plateau. This plateau is
expected given the circular nature of the algorithm: the learner
inspects the region of the solution space that is typical of solutions
already found. We see in this phenomenon a conceptual demarcation of
the problem space: a coherent set of concepts or problems solving
techniques can define by their power the extent of problems a problem
solver is willing to consider. One might hope, however, that an
intelligent learner would recognize that such a demarcation has
occured and attempt to find another class of concepts for the
remaining unsolved tasks.

Another aspect of conceptual learning that we see present in this
algorithm's behavior is the importance of a curriculum. By this we do
not mean an explicit ordering of the problems in the task set, but
rather that a set of sufficiently simple problems must act as stepping
stones to the acquisition of complex representations.
